---
title: "02 - QC, filtering, clustering and batch correction"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
params:
  h5_dir: "../data/cellbender_h5"
  sample_table: "../sample_tables/sample_table_final12.txt"
  output_rds: "../data/intermediate/sobj_batch.rds"
  markers_csv: "../data/intermediate/cluster_markers.csv"
  markers_cca_csv: "../data/intermediate/cluster_markers_cca.csv"
date: "`r Sys.Date()`"
---

Starting point: CellBender-denoised per-sample count matrices (see
[`01_upstream_processing.md`](01_upstream_processing.md)). This Rmd loads
them, applies QC filtering, runs the standard Seurat clustering workflow,
then batch-corrects across donors (`IntegrateLayers`, CCA) and re-clusters
on the integrated embedding. Its output (`params$output_rds`) is the input
to [`03_annotation_label_transfer.Rmd`](03_annotation_label_transfer.Rmd).

By default this renders the final 12-sample analysis (`sample_table_final12.txt`,
donor 3 already excluded) that every figure in this repo is built on.
SFig12's donor-3-outlier diagnostic (panels A/E) does **not** need this Rmd
re-run — it uses precomputed pseudobulk count tables already vendored in
[`../figures/SFig12_data/`](../figures/SFig12_data/). Re-running this Rmd
with `params$sample_table` pointed at `../sample_tables/sample_table_full16.txt`
(and `params$output_rds` changed to avoid overwriting the default output) is
only needed if you want to rebuild that diagnostic from scratch instead —
see [`../docs/donor3_exclusion.md`](../docs/donor3_exclusion.md).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(Seurat)
  library(tidyverse)
  library(hdf5r)
  library(DT)
})

options(future.globals.maxSize = 48 * 1024 * 1024^2)
source("../R/io_helpers.R")

dir.create(dirname(params$output_rds), recursive = TRUE, showWarnings = FALSE)

require_files(c(sample_table = params$sample_table, h5_dir = params$h5_dir))
sample_table <- read.delim(params$sample_table, header = TRUE)
sample_table
```

## Load CellBender-denoised matrices and merge

```{r load merge}
h5_files <- list.files(params$h5_dir, pattern = "BC.*filtered_seurat.*h5$",
                        recursive = TRUE, full.names = TRUE)
names(h5_files) <- sub("_.*", "", sub(".*BC", "BC", h5_files))
h5_files <- h5_files[names(h5_files) %in% sample_table$barcode]
stopifnot("no CellBender h5 files found for any sample in params$sample_table - check params$h5_dir" = length(h5_files) > 0)

sobjs <- lapply(names(h5_files), function(bc) {
  Read10X_h5(h5_files[[bc]]) %>%
    CreateSeuratObject(project = bc, assay = "RNA", min.cells = 1, min.features = 20)
})

sobj_cb <- merge(sobjs[[1]], sobjs[2:length(sobjs)], add.cell.ids = names(h5_files), project = "A4869_Foti") %>%
  JoinLayers(assay = "RNA")
```

## QC metrics and filtering

Cutoffs: `pct.mito < 10`, `500 <= nFeature_RNA < 6000`, genes detected in >=1 cell.

```{r qc metrics}
md <- data.frame(barcode = sub("_.*", "", Cells(sobj_cb)), cells = Cells(sobj_cb)) %>%
  left_join(sample_table, by = "barcode") %>%
  column_to_rownames("cells")
sobj_cb <- AddMetaData(sobj_cb, metadata = md)
sobj_cb[["pct.mito"]] <- PercentageFeatureSet(sobj_cb, pattern = "^MT-")

ggplot(sobj_cb[[]] %>% select(sample, nCount_RNA, nFeature_RNA, pct.mito) %>%
         gather(metric, value, -sample),
       aes(x = sample, y = value, fill = sample)) +
  geom_boxplot(outlier.size = .5) +
  facet_wrap(~metric, ncol = 4, scales = "free_y") +
  theme_classic() +
  theme(axis.text.x = element_blank(), legend.position = "bottom")
```

```{r filter}
mito_cutoff <- 10
nFeature_cutoff <- 6000

sobj <- CreateSeuratObject(counts = GetAssayData(sobj_cb, assay = "RNA", layer = "counts"),
                            min.cells = 1, min.features = 500)
sobj[["pct.mito"]] <- PercentageFeatureSet(sobj_cb, pattern = "^MT-")
sobj <- subset(sobj, (pct.mito < mito_cutoff) & (nFeature_RNA < nFeature_cutoff))
sobj <- AddMetaData(sobj, metadata = md[Cells(sobj), , drop = FALSE])

sobj[[]] %>%
  group_by(sample) %>%
  summarise(ncells = n(), nGene = median(nFeature_RNA), nUMI = median(nCount_RNA), pct.mito = mean(pct.mito))
```

## Standard Seurat workflow (per-sample, pre-batch-correction)

```{r cluster}
s.genes <- cc.genes$s.genes[cc.genes$s.genes %in% rownames(sobj)]
g2m.genes <- cc.genes$g2m.genes[cc.genes$g2m.genes %in% rownames(sobj)]

sobj <- sobj %>%
  NormalizeData(normalization.method = "LogNormalize", scale.factor = 10000, verbose = FALSE) %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000, verbose = FALSE) %>%
  CellCycleScoring(s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE) %>%
  ScaleData(vars.to.regress = c("S.Score", "G2M.Score"), verbose = FALSE) %>%
  RunPCA(features = VariableFeatures(.), verbose = FALSE) %>%
  FindNeighbors(dims = 1:15, verbose = FALSE) %>%
  RunUMAP(dims = 1:15, verbose = FALSE) %>%
  FindClusters(resolution = .6, verbose = FALSE)

ElbowPlot(sobj)
DimPlot(sobj, group.by = "sample") | DimPlot(sobj, group.by = "treatment") | DimPlot(sobj, group.by = "batch")
DimPlot(sobj, label = TRUE, repel = TRUE)
```

```{r markers, fig.width=8, fig.height=10}
Idents(sobj) <- "seurat_clusters"
markers <- FindAllMarkers(sobj, only.pos = TRUE, min.pct = 0.25, verbose = FALSE)
write.csv(markers, params$markers_csv, row.names = FALSE)

top5 <- markers %>% group_by(cluster) %>% slice_max(avg_log2FC, n = 5, with_ties = FALSE)
DotPlot(sobj, features = unique(top5$gene)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(x = "", y = "") + coord_flip() + ggtitle("top5 markers per cluster")

markers %>%
  mutate_if(is.numeric, signif, 2) %>%
  DT::datatable(rownames = FALSE, extensions = "Buttons", options = list(dom = "frtBip", buttons = c("csv")))
```

## Batch correction (CCA integration across donors) and re-clustering

```{r integrate}
sobj[["RNA"]] <- split(sobj[["RNA"]], f = sobj$batch)
sobj <- IntegrateLayers(object = sobj, method = CCAIntegration,
                         orig.reduction = "pca", new.reduction = "integrated.cca", verbose = FALSE)
sobj <- FindNeighbors(sobj, reduction = "integrated.cca", dims = 1:30)
sobj <- FindClusters(sobj, resolution = .8, cluster.name = "cca_clusters")
sobj <- RunUMAP(sobj, reduction = "integrated.cca", dims = 1:30, reduction.name = "iUMAPcca")
sobj <- JoinLayers(sobj)
```

```{r int dims, fig.width=15, fig.height=6}
DimPlot(sobj, group.by = "cca_clusters", reduction = "iUMAPcca", label = TRUE, repel = TRUE) |
  DimPlot(sobj, group.by = "batch", reduction = "iUMAPcca", label = TRUE, repel = TRUE)
DimPlot(sobj, group.by = "sample", reduction = "iUMAPcca", label = TRUE, repel = TRUE) |
  DimPlot(sobj, group.by = "treatment", reduction = "iUMAPcca", label = TRUE, repel = TRUE)
```

```{r markers cca, fig.width=8, fig.height=15}
DefaultAssay(sobj) <- "RNA"
Idents(sobj) <- "cca_clusters"
markers_cca <- FindAllMarkers(sobj, only.pos = TRUE, min.pct = 0.25, verbose = FALSE)
write.csv(markers_cca, params$markers_cca_csv, row.names = FALSE)

top5 <- markers_cca %>% group_by(cluster) %>% slice_max(avg_log2FC, n = 5, with_ties = FALSE)
DotPlot(sobj, features = unique(top5$gene)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(x = "", y = "") + coord_flip() + ggtitle("top5 markers per cca cluster")

markers_cca %>%
  mutate_if(is.numeric, signif, 2) %>%
  DT::datatable(rownames = FALSE, extensions = "Buttons", options = list(dom = "frtBip", buttons = c("csv")))
```

```{r save}
saveRDS(sobj, params$output_rds)
```

```{r session info}
sessionInfo()
```
