---
title: "Figure 5"
output:
  html_document:
    df_print: paged
    code_folding: hide
params:
  pbmc_rds: "../data/intermediate/20250919_pbmc_reproduced.rds"
  de_dir: "../data/intermediate/de_results"
  output_dir: "../output/Fig5"
date: "`r Sys.Date()`"
---

Reproduces Figure 5, panels A-H (the scRNA-seq/pseudobulk panels; panels I-J
are an in vivo rat peritonitis experiment, out of scope for this repo).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(tmod)
})
source("../R/plot_theme.R")
source("../R/io_helpers.R")
source("../R/panel_helpers.R")

require_files(c(pbmc_rds = params$pbmc_rds))
sobj <- readRDS(params$pbmc_rds)
de <- load_de_results(params$de_dir)
res_by_contrast <- de$by_contrast
long_res <- de$long
```

## A - UMAP by cell type

```{r panel A, fig.width=4, fig.height=3}
panel_a <- DimPlot(sobj, group.by = "predicted.celltype.l1", reduction = "umap",
                    label = TRUE, repel = TRUE, label.size = 3) +
  NoLegend() + ggtitle("") +
  theme(axis.text = element_text(size = panel_theme_size), axis.title = element_text(size = panel_theme_size))
panel_a

data_a <- sobj[[]] %>% select(sample, treatment, predicted.celltype.l1) %>%
  cbind(Embeddings(sobj, "umap")) %>% tibble::rownames_to_column("cell")
```

## B - DE gene counts per cell type, all contrasts

```{r panel B, fig.width=8, fig.height=3}
panel_b <- deg_count_bar(long_res)
panel_b

data_b <- long_res %>% filter(!grepl("DEPRECATED", gene_name), !grepl("^ENSG", gene_name), padj < 0.05, cell_type != "all") %>%
  count(cell_type, contrast, name = "n_de_genes")
```

## C-E - Monocyte volcano plots, Cl_vs_UM / I_vs_UM / Cl_vs_I

```{r panel CDE, fig.width=8, fig.height=3}
# volcano_panel() hides its own legend by default (legend.position="none",
# shared with SFig13 B's use of the same function) since the paper shows one
# shared padj<0.05 legend for the whole 3-panel row instead of one per
# panel. Turn panel_c's legend back on and let plot_layout(guides="collect")
# gather it once for all three - panel_d/e are left as-is (no legend) so
# nothing gets duplicated.
panel_c <- volcano_panel(res_by_contrast, "Cl_vs_UM", "Mono", max_overlaps = 30) +
  theme(legend.position = "right",
        legend.text=element_text(size=8),
        legend.title=element_text(size=8)) +
  guides(colour = guide_legend(title = "padj < 0.05"))
panel_d <- volcano_panel(res_by_contrast, "I_vs_UM", "Mono", max_overlaps = 30)
panel_e <- volcano_panel(res_by_contrast, "Cl_vs_I", "Mono", max_overlaps = 80)
panel_cde <- (panel_c | panel_d | panel_e) + plot_layout(guides = "collect")
panel_cde

mono_sig <- function(contrast) res_by_contrast[[contrast]] %>% filter(!is.na(padj), baseMean > 5, cell_type == "Mono")
data_c <- mono_sig("Cl_vs_UM")
data_d <- mono_sig("I_vs_UM")
data_e <- mono_sig("Cl_vs_I")
```

## F - tmod enrichment, monocytes, Cl_vs_UM / I_vs_UM / Cl_vs_I

```{r panel F, fig.width=8, fig.height=6}
data(tmod)

f_contrasts <- c("Cl_vs_UM", "I_vs_UM", "Cl_vs_I")
d <- long_res %>% filter(baseMean > 5, cell_type == "Mono", contrast %in% f_contrasts)
l <- split(d, factor(contrast_labels_2lines[d$contrast], levels = contrast_labels_2lines[f_contrasts]))

genes <- lapply(l, function(x) x %>% filter(!is.na(padj)) %>% arrange(pvalue))
lgenes <- lapply(genes, function(x) x$gene_name)
res.tmod <- lapply(lgenes, tmodCERNOtest, mset = tmod)
sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange, pval = x$padj,
                                                  lfc.thr = 0, pval.thr = 0.05, mset = tmod))
names(sgenes) <- names(l)

# a handful of near-duplicate/low-informative BTM modules excluded, as in the original analysis
filter_ids <- c("LI.S11", "LI.M89.1", "LI.M61.2", "LI.M7.3", "LI.M7.0", "DC.M4.1", "LI.M25", "DC.M3.6", "LI.M7.2")
res.tmod.f <- lapply(res.tmod, function(x) x %>% filter(!ID %in% filter_ids, N1 > 5, N1 < 200))

panel_f <- ggPanelplot(res.tmod.f, sgenes = sgenes, auc_thr = 0, q_thr = 1,
                        filter_row_q = 5e-4, filter_row_auc = 0.7, q_cutoff = 1e-12,
                        label_angle = 0, add_ids = FALSE, mset = tmod, cluster = TRUE) +
  theme(axis.text = element_text(size = panel_theme_size), strip.text = element_text(size = panel_theme_size),
        legend.text = element_text(size = panel_theme_size), legend.title = element_text(size = panel_theme_size),
        axis.title = element_text(size = panel_theme_size), strip.text.x = element_text(hjust = 0),
        axis.text.x = element_text(angle = 0), panel.background = element_blank(),
        legend.key.size = unit(.1, "in")) +
  guides(alpha = guide_legend(ncol = 1))
panel_f

# Mirrors ggPanelplot()'s filter_row_q/filter_row_auc row selection above (a
# module is kept, across every contrast column, if it passes the threshold
# in at least one contrast) - res.tmod.f itself is pre-ggPanelplot data, so
# exporting it directly would include every tested module, not just the ones
# that actually pass filter_row_q/filter_row_auc and get drawn as heatmap
# rows.
combined_f <- Map(cbind, res.tmod.f, contrast = names(res.tmod.f)) %>% bind_rows()
keep_ids_f <- combined_f %>% filter(adj.P.Val < 5e-4, AUC > 0.7) %>% pull(ID) %>% unique()
data_f <- combined_f %>% filter(ID %in% keep_ids_f)
```

## G-H - Monocyte chemokine-cluster scatter, ClY21-HNP1 / IY16-HNP1 vs vehicle

Genes highlighted (red dots + labels) belong to the manuscript's "chemokine
cluster (II)" (see Results text): ANXA1, CXCL5, CCL7, CCR1, CCL2.

```{r panel GH, fig.width=3, fig.height=3}
chemokine_cluster_ii <- c("ANXA1", "CXCL5", "CCL7", "CCR1", "CCL2")

panel_g <- concordance_scatter(res_by_contrast, "UM_vs_X", "Cl_vs_X", "Mono",
                                style = "highlight", highlight_genes = chemokine_cluster_ii,
                                title = "Monocytes", xlab = "log2FC HNP1 vs vehicle", ylab = "log2FC ClY21-HNP1 vs vehicle")
panel_h <- concordance_scatter(res_by_contrast, "UM_vs_X", "I_vs_X", "Mono",
                                style = "highlight", highlight_genes = chemokine_cluster_ii,
                                title = "Monocytes", xlab = "log2FC HNP1 vs vehicle", ylab = "log2FC IY16-HNP1 vs vehicle")
panel_g | panel_h

scatter_data <- function(contrast_x, contrast_y) {
  # Matches concordance_scatter()'s own filtering (R/panel_helpers.R) so the
  # exported table is exactly the set of genes shown in the plot, not every
  # gene DESeq2 tested - drop deprecated/Ensembl-ID fallback names, and keep
  # only genes significant in at least one of the two contrasts.
  inner_join(
    res_by_contrast[[contrast_x]] %>% filter(cell_type == "Mono") %>% select(gene_name, log2FC_x = log2FoldChange, padj_x = padj),
    res_by_contrast[[contrast_y]] %>% filter(cell_type == "Mono") %>% select(gene_name, log2FC_y = log2FoldChange, padj_y = padj),
    by = "gene_name"
  ) %>%
    filter(!grepl("DEPRECATED", gene_name), !grepl("^ENSG", gene_name)) %>%
    filter((!is.na(padj_x) & padj_x < 0.05) | (!is.na(padj_y) & padj_y < 0.05)) %>%
    mutate(chemokine_cluster_ii = gene_name %in% chemokine_cluster_ii)
}
data_g <- scatter_data("UM_vs_X", "Cl_vs_X")
data_h <- scatter_data("UM_vs_X", "I_vs_X")
```

## Assembly

```{r assembly, fig.width=8, fig.height=11}
fig5 <- ((panel_a) + (panel_b) + plot_layout(widths=c(0.5,1))) /
  wrap_elements(panel_cde) /
  (wrap_elements(panel_f)) /
  (wrap_elements(panel_g) + wrap_elements(panel_h) + plot_spacer()) +
  plot_layout(heights = c(1.2, 1.5, 2,1.5)) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 8, face = "bold", hjust = -.5, vjust = 0))
fig5
```

```{r save}
save_figure_pdf(fig5, file.path(params$output_dir, "Fig5.pdf"), width = 8, height = 11)
write_panel_xlsx(
  list(A_umap_celltype = data_a, B_deg_counts = data_b, C_volcano_ClvsUM_Mono = data_c,
       D_volcano_IvsUM_Mono = data_d, E_volcano_ClvsI_Mono = data_e, F_tmod_Mono = data_f,
       G_scatter_ClvsX_Mono = data_g, H_scatter_IvsX_Mono = data_h),
  file.path(params$output_dir, "Fig5_data.xlsx")
)
```

```{r session info}
sessionInfo()
```
