---
title: "01 - Pseudobulk count tables"
output:
  html_document:
    df_print: paged
    code_folding: hide
params:
  annotated_rds: "../data/intermediate/sobj_pbmc_annotated.rds"
  sample_table: "../sample_tables/sample_table_final12.txt"
  output_counts_l1: "../data/intermediate/counts_l1.tsv"
  output_counts_l2: "../data/intermediate/counts_l2.tsv"
date: "`r Sys.Date()`"
---

Sums raw counts per (sample x cell type) to build pseudobulk count matrices,
at both `predicted.celltype.l1` and `predicted.celltype.l2` granularity, plus
an `all_<sample>` pseudobulk per sample (all cells, any type). Input is the
annotated object from
[`../preprocessing/03_annotation_label_transfer.Rmd`](../preprocessing/03_annotation_label_transfer.Rmd)
(`sobj_pbmc_annotated.rds`, pre-reference-UMAP-merge; NOT the final
`20250919_pbmc.rds`). Output feeds
[`02_deseq2_tmod.Rmd`](02_deseq2_tmod.Rmd) (l1 only) and the PCA QC plots below.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(Seurat)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DESeq2)
})
source("../R/io_helpers.R")
dir.create(dirname(params$output_counts_l1), recursive = TRUE, showWarnings = FALSE)

require_files(c(annotated_rds = params$annotated_rds, sample_table = params$sample_table))
sobj <- readRDS(params$annotated_rds)
sample_table <- read.delim(params$sample_table)
```

```{r umap}
DimPlot(sobj, group.by = "predicted.celltype.l1", reduction = "iUMAPcca", label = TRUE, repel = TRUE) + NoLegend()
```

```{r pseudobulk fn}
pseudobulk <- function(sobj, celltype_col) {
  count_mtx <- GetAssayData(sobj, layer = "counts")
  cell_types <- sobj[[]] %>% pull(!!celltype_col) %>% unique() %>% as.character()
  samples <- unique(sobj[[]]$sample)

  expr <- list()
  for (ct in cell_types) {
    for (s in samples) {
      cells <- Cells(sobj)[sobj@meta.data$sample == s & sobj@meta.data[[celltype_col]] == ct]
      cells <- cells[!is.na(cells)]
      if (length(cells) > 1) expr[[paste0(ct, "_", s)]] <- rowSums(count_mtx[, cells])
    }
  }
  for (s in samples) {
    cells <- Cells(sobj)[sobj@meta.data$sample == s]
    cells <- cells[!is.na(cells)]
    expr[[paste0("all_", s)]] <- rowSums(count_mtx[, cells])
  }
  names(expr) <- gsub(" ", "_", names(expr))

  colData <- data.frame(sample_name = factor(names(expr)),
                         cell_type = factor(sub("_s.*", "", names(expr))),
                         sample = factor(sub(".*_s", "s", names(expr)))) %>%
    left_join(sample_table, by = "sample") %>%
    mutate(batch = factor(batch))

  counts <- do.call(cbind, expr) %>% as.data.frame() %>% tibble::rownames_to_column("gene_name")
  list(counts = counts, colData = colData)
}

flag_low_depth <- function(counts, sample_table, min_reads) {
  counts_mat <- counts %>% tibble::column_to_rownames("gene_name")
  lengths <- colSums(counts_mat)
  df <- data.frame(sample_name = names(lengths), n_reads = lengths) %>%
    mutate(sample = sub(".*_s", "s", sample_name), celltype = sub("_s.*", "", sample_name)) %>%
    left_join(sample_table, by = "sample") %>%
    mutate(batch = factor(batch))
  print(
    ggplot(df, aes(group, n_reads, colour = group)) +
      geom_point(position = position_dodge(width = .5), size = 2) +
      facet_wrap(~celltype, scales = "free") +
      theme(axis.text.x = element_blank())
  )
  unique(df[df$n_reads < min_reads, "celltype"])
}

plotPCA_per_celltype <- function(counts, colData, celltype) {
  cd <- filter(colData, cell_type == celltype)
  cnt <- unique(cd$sample_name)
  counts_mat <- counts %>% tibble::column_to_rownames("gene_name")
  dds <- DESeqDataSetFromMatrix(countData = counts_mat[, cnt], colData = cd, design = ~1) %>% DESeq()
  rld <- vst(dds)
  for (cov in c("sample", "group", "batch")) {
    print(plotPCA(rld, intgroup = cov) + guides(colour = guide_legend(title = cov)) + ggtitle(celltype))
  }
}
```

## Granularity level 1 (used by DE + all figures)

```{r l1}
res_l1 <- pseudobulk(sobj, "predicted.celltype.l1")
write.table(res_l1$counts, params$output_counts_l1, quote = FALSE, sep = "\t", row.names = FALSE)
exclude_l1 <- flag_low_depth(res_l1$counts, sample_table, min_reads = 5e5)
exclude_l1
```

```{r l1 pca, results='hide'}
invisible(lapply(c("all", "Mono"), plotPCA_per_celltype, counts = res_l1$counts, colData = res_l1$colData))
```

## Granularity level 2 (QC only; not used by downstream DE/figures)

```{r l2}
res_l2 <- pseudobulk(sobj, "predicted.celltype.l2")
write.table(res_l2$counts, params$output_counts_l2, quote = FALSE, sep = "\t", row.names = FALSE)
exclude_l2 <- flag_low_depth(res_l2$counts, sample_table, min_reads = 2e5)
exclude_l2
```

```{r l2 pca, results='hide'}
invisible(lapply(c("all", "CD14_Mono"), plotPCA_per_celltype, counts = res_l2$counts, colData = res_l2$colData))
```

```{r session info}
sessionInfo()
```
