---
title: "02 - Pseudobulk DESeq2 + tmod gene set enrichment"
output:
  html_document:
    df_print: paged
    code_folding: hide
params:
  counts_l1: "../data/intermediate/counts_l1.tsv"
  sample_table: "../sample_tables/sample_table_final12.txt"
  shrinkage_type: "normal"
  output_dir: "../data/intermediate/de_results"
date: "`r Sys.Date()`"
---

Pseudobulk differential expression per cell type (design `~batch + group`),
6 pairwise contrasts among the 4 treatments (`X` = vehicle, `UM` = HNP1,
`Cl` = ClY21-HNP1, `I` = IY16-HNP1), followed by tmod gene set enrichment
(BTM modules, the built-in `tmod` set — this is the only gene set collection
used in the manuscript; no MSigDB/Reactome/KEGG/GO enrichment was part of
the final analysis). Per the manuscript Methods: *"We extracted pairwise
comparisons between different groups from the same results object, to
consider the total variance from all samples in the experiment."*

Input is `counts_l1.tsv` from
[`01_pseudobulk_count_tables.Rmd`](01_pseudobulk_count_tables.Rmd). Output
(`params$output_dir/results_<contrast>.tsv`, one of 6 files) is what every
figure Rmd reads to build its DE-based panels.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggrepel)
  library(DESeq2)
  library(tmod)
  library(DT)
})
source("../R/io_helpers.R")
dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)

require_files(c(counts_l1 = params$counts_l1, sample_table = params$sample_table))
sample_table <- read.delim(params$sample_table)
counts <- read.delim(params$counts_l1, header = TRUE) %>% tibble::column_to_rownames("gene_name")

selected_contrasts <- c("UM_vs_X", "Cl_vs_X", "I_vs_X", "Cl_vs_UM", "I_vs_UM", "Cl_vs_I")
```

```{r coldata}
colData <- data.frame(sample_name = factor(colnames(counts)),
                       cell_type = factor(sub("_s.*", "", colnames(counts))),
                       sample = factor(sub(".*_s", "s", colnames(counts)))) %>%
  left_join(sample_table, by = "sample") %>%
  mutate(batch = factor(batch)) %>%
  filter(cell_type != "other")

selected_cell_types <- unique(colData$cell_type)
selected_contrasts
```

## DESeq2, all cell types x all contrasts

```{r deseq2}
res <- list()
for (celltype in selected_cell_types) {
  coldata_df <- filter(colData, cell_type == celltype)
  counts_df <- subset(counts, select = as.character(coldata_df$sample_name))
  if (sum(colSums(counts_df)) == 0) next

  dds <- DESeqDataSetFromMatrix(countData = as.matrix(counts_df), colData = coldata_df, design = ~batch + group)

  sub_res <- list()
  for (comparison in selected_contrasts) {
    perturbation <- sub("_vs_.*", "", comparison)
    reference <- sub(".*_vs_", "", comparison)
    dds$group <- relevel(dds$group, ref = reference)
    dds <- if (celltype %in% c("all", "all_cells")) estimateSizeFactors(dds) else estimateSizeFactors(dds, type = "poscounts")
    dds <- DESeq(dds)

    sub_res[[comparison]] <- (if (params$shrinkage_type == "apeglm") {
      lfcShrink(dds, coef = paste0("group_", comparison), type = params$shrinkage_type, format = "DataFrame")
    } else {
      lfcShrink(dds, contrast = c("group", perturbation, reference), type = params$shrinkage_type, format = "DataFrame")
    }) %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_name") %>%
      mutate(cell_type = celltype, contrast = comparison)
  }
  res[[celltype]] <- do.call(rbind, sub_res) %>% tibble::remove_rownames()
}
res_df <- do.call(rbind, res) %>% tibble::remove_rownames()
```

```{r save results}
for (contrast_name in unique(res_df$contrast)) {
  write.table(filter(res_df, contrast == contrast_name),
              file.path(params$output_dir, paste0("results_", contrast_name, ".tsv")),
              quote = FALSE, sep = "\t", row.names = FALSE)
}
```

## QC: volcano plots per contrast, all cell types

```{r volcano, fig.height=18, fig.width=10}
for (comparison in unique(res_df$contrast)) {
  print(
    res_df %>%
      filter(!is.na(padj), contrast == comparison) %>%
      ggplot(aes(x = log2FoldChange, y = -log10(pvalue), color = padj < .05)) +
      geom_point(size = .5, shape = 1) +
      geom_label_repel(data = filter(res_df, contrast == comparison, padj < .05),
                        aes(label = gene_name), segment.size = .25, segment.alpha = .5, size = 3,
                        color = "black", max.overlaps = 50, label.padding = 0.05, label.size = NA) +
      scale_color_manual(values = c("black", "red")) +
      theme_classic() + theme(legend.position = "none", strip.background = element_blank()) +
      ggtitle(comparison) + facet_wrap(~cell_type, ncol = 2, scales = "free")
  )
}
```

```{r show results table}
res_df %>%
  filter(padj < .05) %>%
  mutate_if(is.numeric, signif, 2) %>%
  DT::datatable(rownames = FALSE, extensions = "Buttons", options = list(dom = "frtBip", buttons = c("csv")))
```

## QC: tmod (BTM) gene set enrichment per contrast, all cell types

Same thresholding used for the manuscript's enrichment figures (Fig 5F,
SFig13): `AUC > 0.70`. This QC pass uses a lenient q cutoff just to preview
which modules turn up; each figure Rmd applies the exact per-panel q
threshold stated in its legend.

```{r tmod, fig.width=15, fig.height=10}
data(tmod)
res_by_contrast <- split(res_df, res_df$contrast)

enrich_panel <- function(contrast, q_filter = 1e-4, auc_filter = 0.70) {
  l <- split(res_by_contrast[[contrast]], res_by_contrast[[contrast]]$cell_type)
  genes <- lapply(l, function(x) x %>% filter(!is.na(padj)) %>% arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = tmod)

  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange, pval = x$padj,
                                                    lfc.thr = 0, pval.thr = 0.05, mset = tmod))
  names(sgenes) <- names(l)

  keep <- names(res.tmod)[sapply(res.tmod, function(x) nrow(filter(x, adj.P.Val < q_filter, AUC > auc_filter)) > 0)]
  relevant <- res.tmod[keep]
  relevant <- Map(cbind, relevant, cell_type = lapply(names(relevant), function(x) rep(x, nrow(relevant[[x]]))))
  if (length(relevant) > 0) {
    print(ggPanelplot(relevant, sgenes = sgenes[names(relevant)], auc_thr = 0, q_thr = 1,
                       filter_row_q = q_filter, filter_row_auc = auc_filter, q_cutoff = 1e-12) + ggtitle(contrast))
  }
  do.call(rbind, relevant)
}

res_tmod <- lapply(names(res_by_contrast), enrich_panel)
names(res_tmod) <- names(res_by_contrast)
res_tmod <- res_tmod[sapply(res_tmod, length) > 0]
```

```{r tmod table}
res_tmod_df <- Map(cbind, res_tmod, contrast = lapply(names(res_tmod), function(x) rep(x, nrow(res_tmod[[x]])))) %>%
  bind_rows()
res_tmod_df %>%
  mutate_if(is.numeric, signif, 2) %>%
  DT::datatable(rownames = FALSE, extensions = "Buttons", options = list(dom = "frtBip", buttons = c("csv")))
```

```{r session info}
sessionInfo()
```
