---
output: 
  html_document:
    df_print: paged
    code_folding: hide
params:
  title: "figures 190925"
  shrinkage_type: "normal"
title: "`r params$title`"
---



```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache=FALSE, cache.lazy=FALSE, message=FALSE, warning=FALSE)

suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(gtools))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(tmod))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(patchwork))



options(future.globals.maxSize = 12 * 1024 * 1024^2)


path <- "/data/cephfs-1/work/groups/cubi/projects/2022-04-27_CSCS_Miha/foti_redone/downstream_analysis/"

results_dir <- "20250919_results"


# environment R-fixed-biomart-deseq

sample_table <- read.delim(file.path(path,"sample_table_20250919.txt"))

plot_names <- structure(c("HNP1", "ClY21-HNP1", "IY16-HNP1"), names=c("UM", "Cl", "I"))

```

```{r object}
sobj <- readRDS(file.path(path,"20250919_results/20250919_pbmc.rds"))

```

For this DE analysis, we used the following parameters:

```{r print parameters}

pars <- as.data.frame(do.call("rbind", params))
colnames(pars) <- "parameters_used"
pars

```


```{r f5a,fig.width=4,fig.height=3}


f5a <- DimPlot(sobj, group.by="predicted.celltype.l1", reduction="umap",label=T, repel=T,label.size = 3)+
  NoLegend()+
  ggtitle("")+
  theme(axis.text=element_text(size=8),
        axis.title=element_text(size=8))
f5a
# DimPlot(sobj, group.by="predicted.celltype.l2", reduction="umap",label=T, repel=T)+NoLegend()

```

```{r s6c,fig.width=4,fig.height=3}
Idents(sobj) <- "predicted.celltype.l1"

features=c("PTPRC", "CD8A", "CD4","GZMK", "CCL5", "S100A4", "ANXA1", "CCR7", "ISG15", "CD3D", "CD3E","CD19", "CSF1R", "S100A8","MS4A2","CD14","CD1C","NCAM1")

library(RColorBrewer)
s6c <- DotPlot(sobj, features = features,dot.scale = 3)+ 
  scale_colour_gradientn(colours =rev(brewer.pal(n = 10, name = "RdBu"))) + 
  theme(text = element_text(family = "Arial"),
        axis.text.x=element_text(angle=90,hjust=1,vjust=.5, size=8),
        axis.text.y=element_text(size=8),
        strip.text.x = element_text(size = 8, angle=90),#for facet wrap
        strip.placement = "down",
        axis.title = element_text(size=8,face="bold"),
        axis.title.y.right = element_text(size = 8),
        legend.text=element_text(size=8),
        legend.position = "top",
        legend.title=element_text(size=8),
        legend.key.width = unit(.05,"in"),
        legend.key.height = unit(.15,"in"),
        axis.line = element_line(linewidth=1)) +
   guides(colour=guide_legend(ncol=2),
          size=guide_legend(ncol=2))+
labs(x = c(), y = c())

s6c
```

 
```{r, fig.width=8,fig.height=12}

files <- list.files("/data/cephfs-1/work/groups/cubi/projects/2022-04-27_CSCS_Miha/foti_redone/downstream_analysis/20250919_results", pattern = "results.*tsv", full.names = T)

res <- lapply(files, read.delim, header=T)
names(res) <- sub("_1909.*","",sub(".*list_","",files))

```
 


```{r load counts}

counts <- read.delim(file.path(path, "20250919_results","counts_l1_190925.tsv"))

```

```{r cell numbers layer1,fig.height=10,fig.width=12}

counts <- counts %>%
  tibble::column_to_rownames("gene_name")

lengths <- counts %>% 
  colSums()

df_lengths <- data.frame(sample = names(lengths),
                         n_reads = lengths) %>%
  mutate(sample_name=sample) %>%
  mutate(sample=sub(".*_s","s",sample_name), celltype=sub("_s.*","",sample_name)) %>% 
  left_join(sample_table) %>%
  mutate(batch=factor(batch))
  

# ggplot(df_lengths, aes(group,n_reads,colour=group))+
#   geom_point(position=position_dodge(width=.5), size=2) +
#   facet_wrap(~celltype, scales="free")+
#   theme(axis.text.x=element_blank())


```

The following cell types might have too little reads to do DE pseudobulk analysis on.

```{r}

exclude <- df_lengths[which(df_lengths$n_reads<5e5),"celltype"] %>% unique
exclude
```

Plot PCA for each cell type and covariates.

```{r plot PCA, results=F, fig.width=6,fig.height=6}

colData<-data.frame(sample_name=factor(names(counts)),
               cell_type=factor(sub("_s.*","",names(counts))),
               sample=factor(sub(".*_s","s", names(counts)))) %>%
  left_join(sample_table) %>%
  mutate(batch=factor(batch))

celltypes <- colData %>%
  filter(!(cell_type %in% exclude)) %>%
  pull(cell_type) %>%
  unique() 

plotPCA_per_celltype <- function(celltype){
  # celltype <- "all"
  cd <- filter(colData, cell_type==celltype)
  cnt <- cd %>%
    pull(sample_name) %>% unique()
  
  dds <- DESeqDataSetFromMatrix(countData=counts[,cnt],
                                colData=cd,
                                design=~1) %>% 
    DESeq()
  rld <- vst(dds)
  
  for (cov in c("sample", "group", "batch")){
    p <- plotPCA(rld, intgroup=cov) + 
      guides(colour=guide_legend(title=cov)) +
      ggtitle(celltype)
    print(p)
  }
  
}

# lapply(c("all" ), plotPCA_per_celltype )

cd <- filter(colData, cell_type=="all")
cnt <- cd %>%
  pull(sample_name) %>% unique()

dds <- DESeqDataSetFromMatrix(countData=counts[,cnt],
                              colData=cd,
                              design=~1) %>% 
  DESeq()
rld <- vst(dds)

p <- plotPCA(rld, intgroup=c("sample", "batch"), returnData=T)  %>%
  ggplot(aes(PC1, PC2, color=sample, shape=batch)) +geom_point(size=3) 
print(p)

p <- plotPCA(rld, intgroup=c("group", "batch"), returnData=T)  %>%
  ggplot(aes(PC1, PC2, color=group.1, shape=batch)) +geom_point(size=3) 
print(p)

``` 

```{r prepare coldata}

sample_name <- sub(".*_s", "s", colnames(counts))
cell_type <- sub("_s.*", "", colnames(counts))

colData<-data.frame(sample_name=factor(colnames(counts)),
               cell_type=factor(cell_type),
               sample=factor(sample_name)) %>%
  left_join(sample_table) %>%
  mutate(batch=factor(batch)) %>%
  filter(cell_type!="other")


               
selected_cell_types <- "all"
  



selected_contrasts <- c("UM_vs_X")

selected_contrasts
```


```{r all contrasts}

celltype <- "Mono"
coldata_df <- colData %>%
  dplyr::filter(cell_type %in% celltype) 

counts_df <- subset(counts, select = as.character(coldata_df$sample_name))
dds <- DESeqDataSetFromMatrix(countData=as.matrix(counts_df),
                              colData=coldata_df,
                              design=~batch + group)
contrasts <- selected_contrasts
sub_res <- list()

comparison<- "UM_vs_X"
perturbation <- sub("_vs_.*", "", comparison)
reference <- sub(".*_vs_", "", comparison)
dds$group <- relevel(dds$group, ref = reference)
dds <- estimateSizeFactors(dds, type="poscounts")
dds <- DESeq(dds)

rlogs <- rlog(dds, blind=F)

data(tmod)
gene_sets <- c("LI.M27.1", "LI.M59", "DC.M1.2") 
genes <- getGenes(gs=gene_sets, mset="all", as.list=T) %>% unlist()
genes_df <- data.frame(gene_set=names(genes), gene=genes)

counts(dds, normalize=T) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene_name") %>%
  filter(gene_name %in% c("CCL7", "CCR1", "CCL2", "OAS2" ,"LY6E", "ISG15", "CCL5", "ISG15", "SOCS2", "OTOF")) %>%
  # filter(gene_name %in% genes_df$gene) %>%
  # filter(rowMeans(.[,-1])>60000) %>%
  tidyr::pivot_longer(-gene_name, names_to="sample_name", values_to="normalized_counts") %>%
  left_join(coldata_df) %>%
  ggplot(aes(group, log2(normalized_counts),colour=group,shape=batch)) +
  geom_point(size=3, position = position_jitterdodge(jitter.width = .1)) +
  facet_wrap(~gene_name, scales="free_y")


# rlogs@assays@data[[1]] %>%
#    as.data.frame() %>%
#   tibble::rownames_to_column("gene_name") %>%
#   filter(gene_name %in% c("CCL7", "CCR1", "CCL2", "OAS2" ,"LY6E", "ISG15", "CCL5", "ISG15", "SOCS2", "OTOF")) %>%
#   # filter(gene_name %in% genes_df$gene) %>%
#   # filter(rowMeans(.[,-1])>60000) %>%
#   tidyr::pivot_longer(-gene_name, names_to="sample_name", values_to="rlog_counts") %>%
#   left_join(coldata_df) %>%
#   ggplot(aes(group, rlog_counts,colour=group,shape=batch)) +
#   geom_point(size=3, position = position_jitterdodge(jitter.width = .1)) +
#   facet_wrap(~gene_name, scales="free_y")
# 
# rld_blind <- switch("vst",
#         "rld"=rlog(dds, blind=TRUE),
#         "vst"=vst(dds, blind=TRUE)
# )
# 
# 
# assay(rld_blind) %>%
#   as.data.frame() %>%
#   tibble::rownames_to_column("gene_name") %>%
#   filter(gene_name %in% c("CCL7", "CCR1", "CCL2", "OAS2" ,"LY6E", "ISG15", "CCL5", "ISG15", "SOCS2", "OTOF")) %>%
#   # filter(gene_name %in% genes_df$gene) %>%
#   # filter(rowMeans(.[,-1])>60000) %>%
#   tidyr::pivot_longer(-gene_name, names_to="sample_name", values_to="rlog_counts") %>%
#   left_join(coldata_df) %>%
#   ggplot(aes(group, rlog_counts,colour=group,shape=batch)) +
#   geom_point(size=3, position = position_jitterdodge(jitter.width = .1)) +
#   facet_wrap(~gene_name, scales="free_y")

```



```{r}

res$Cl_vs_X %>%
  filter(cell_type=="Mono") %>%
  filter(gene_name %in% c("CCL7", "CCR1", "CCL2", "OAS2" ,"LY6E", "ISG15", "CCL5"))
  

relevant_genes <- res$Cl_vs_X %>%
  filter(cell_type=="Mono") %>%
  filter(padj < .05) %>%
  filter(gene_name %in% genes_df$gene) %>%
  pull(gene_name)



```












Volcano plots.

```{r f5c f5d f5e, fig.width=4,fig.height=3}

f5c <- res[["UM_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="Mono") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["UM_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="Mono"),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=80,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("HNP1 vs X", " (Monocytes)"))
print(f5c)

f5d <- res[["Cl_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="Mono") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["Cl_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="Mono"),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=80,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("ClY21-HNP1 vs X", " (Monocytes)"))
print(f5d)



f5e <- res[["I_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="Mono") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["I_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="Mono"),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=80,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("IY16-HNP1 vs X", " (Monocytes)"))
print(f5e)
```

```{r s8a s8b s8c, fig.width=4,fig.height=3}

s8a <- res[["UM_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="DC") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["UM_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="DC") %>%
                     dplyr::filter(!grepl("DEPRECATED",gene_name)),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=50,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("HNP1 vs X", " (Dendritic)"))
print(s8a)

s8b <- res[["Cl_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="DC") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["Cl_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="DC") %>%
                     dplyr::filter(!grepl("DEPRECATED",gene_name)),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=50,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("ClY21-HNP1 vs X", " (Dendritic)"))
print(s8b)



s8c <- res[["I_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="DC") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["I_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="DC") %>%
                     dplyr::filter(!grepl("DEPRECATED",gene_name)),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=50,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("IY16-HNP1 vs X", " (Dendritic)"))
print(s8c)
```

```{r s6d s6e s6f,fig.width=4,fig.height=3}

s6d <- res[["UM_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="B") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["UM_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="B") %>%
                     dplyr::filter(!grepl("DEPRECATED",gene_name)),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=50,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic()+
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("HNP1 vs X", " (B cells)"))
print(s6d)

s6e <- res[["Cl_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="B") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["Cl_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="B") %>%
                     dplyr::filter(!grepl("DEPRECATED",gene_name)),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=50,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("ClY21-HNP1 vs X", " (B cells)"))
print(s6e)



s6f <- res[["I_vs_X"]] %>%                  
  dplyr::filter(!is.na(padj), baseMean>5, cell_type=="B") %>%
  ggplot(aes(x=log2FoldChange,y=-log10(padj),color=padj < .05)) + 
  geom_point(size=.5, shape=1) +
  geom_label_repel(data=res[["I_vs_X"]]  %>%
                     dplyr::filter(padj < .05,baseMean>5, cell_type=="B")%>%
                     dplyr::filter(!grepl("DEPRECATED",gene_name)),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=2.5,color='black',max.overlaps=50,
                   label.padding=0.05,label.size=NA) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color = "grey") +
  coord_cartesian(xlim=c(-4,4))+
  scale_color_manual(values=c('black','red')) +
  theme_classic() +
  theme(strip.background=element_blank(),
    plot.title=element_text(size=8),
    axis.title=element_text(size=8),
    axis.text=element_text(size=8),
    legend.text=element_text(size=8),
    legend.title=element_text(size=8))+
  ggtitle(paste0("IY16-HNP1 vs X", " (B cells)"))
print(s6f)


```



Get enrichments.

```{r f5b, results=T, fig.width=8,fig.height=3}

data(tmod)

plot_cells <- c("B","Mono","DC")

plot_contrasts <- c("HNP1_vs_X\n", "ClY21-HNP1\n_vs_X", "IY16-HNP1\n_vs_X") %>% 
  str_replace_all("_"," ")

rename_contrasts <- structure(c( "ClY21-HNP1\n_vs_IY16-HNP1", "ClY21-HNP1\n_vs_HNP1", "ClY21-HNP1\n_vs_X","IY16-HNP1\n_vs_HNP1", "IY16-HNP1\n_vs_X", "HNP1_vs_X\n" ) %>% str_replace_all("_"," "),
                              names=c("Cl_vs_I" ,"Cl_vs_UM","Cl_vs_X", "I_vs_UM","I_vs_X", "UM_vs_X"))

rcontrast <- lapply(res, function(x) rename_contrasts[x %>% pull(contrast)])
renamed_res <- Map(cbind, res, rcontrast=rcontrast)

names(renamed_res) <- rename_contrasts

l <- renamed_res %>%
  bind_rows() %>%
  filter(baseMean>5) %>%
  filter(cell_type %in% plot_cells, rcontrast %in% plot_contrasts) %>%
  mutate(contrast_cell_type=paste0(rcontrast," ",cell_type)) %>%
  mutate(contrast_cell_type=ifelse(grepl(" B", contrast_cell_type), paste0(contrast_cell_type," cells"),
                                   contrast_cell_type)) %>%
  split(f=.$contrast_cell_type) 

plot_order <- c("HNP1_vs_X\n_B_cells","ClY21-HNP1\n_vs_X_B_cells", "IY16-HNP1\n_vs_X_B_cells", "HNP1_vs_X\n_Mono","ClY21-HNP1\n_vs_X_Mono", "IY16-HNP1\n_vs_X_Mono","HNP1_vs_X\n_DC","ClY21-HNP1\n_vs_X_DC", "IY16-HNP1\n_vs_X_DC" ) %>% str_replace_all("_"," ")

l <- l[plot_order]

  genes <- lapply(l, function(x) x %>%
                    filter(!is.na(padj)) %>%
                    arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = tmod)
  
  
  
  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange,
                                                  pval = x$padj, lfc.thr=0, pval.thr = 0.05, mset = tmod))
  names(sgenes) <- names(l)
  
  ids <- c("LI.S11", "LI.M68", "LI.M67", 
           "LI.M11.0", "DC.M3.4", "LI.M75", 
           "LI.M165")
  
  f5b <- res.tmod %>%
    lapply(function(x) x %>%
             filter(ID %in% ids)) %>%
    ggPanelplot(sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-5,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90,add_ids = F, id_order=rev(ids)) +
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))+
    guides(alpha=guide_legend(ncol=1))
  print(f5b)
  
  
  
```


We extend the pathway analysis to Reactome, Kegg and GO:BP databases.

```{r add pathways, fig.width=25,fig.height=20,eval=T}
library(tmod)
data(tmod)


df2tmod <- function(df, gene_id_col=ncol(df), module_id_col=1, module_title_col=2) {

  require(tmod, quietly=TRUE)

  gene_ids <- df[[gene_id_col]]
  module_ids <- df[[module_id_col]]
  m2g <- tapply(gene_ids, module_ids, list)
  df[[gene_id_col]] <- NULL
  df <- df[!duplicated(df[[module_id_col]]), ]
  colnames(df)[module_id_col] <- "ID"
  colnames(df)[module_title_col] <- "Title"

  makeTmod(modules=df, modules2genes=m2g)
}

df <- as.data.frame(msigdbr::msigdbr(species='Homo sapiens'))
df <- df[ , c("gs_name", "gs_id", "gs_cat", "gs_subcat", "gene_symbol") ]
colnames(df) <- c("Title", "ID", "Category", "Subcategory", "GeneID")
msig <- df2tmod(df[!is.na(df$GeneID),], gene_id_col=ncol(df), module_id_col=2, module_title_col=1)
sel <- (msig$gs$Category %in% c("H"))  | (msig$gs$Subcategory %in% c('CP:REACTOME','GO:BP','CP:KEGG'))

```


```{r other pathways s7a, results=T, fig.width=8,fig.height=3}

plot_cells <- c("B","Mono", "DC")


plot_contrasts <- c("HNP1_vs_X\n", "ClY21-HNP1\n_vs_X", "IY16-HNP1\n_vs_X") %>% 
  str_replace_all("_"," ")

rename_contrasts <- structure(c( "ClY21-HNP1\n_vs_IY16-HNP1", "ClY21-HNP1\n_vs_HNP1", "ClY21-HNP1\n_vs_X","IY16-HNP1\n_vs_HNP1", "IY16-HNP1\n_vs_X", "HNP1_vs_X\n" ) %>% str_replace_all("_"," "),
                              names=c("Cl_vs_I" ,"Cl_vs_UM","Cl_vs_X", "I_vs_UM","I_vs_X", "UM_vs_X"))

rcontrast <- lapply(res, function(x) rename_contrasts[x %>% pull(contrast)])
renamed_res <- Map(cbind, res, rcontrast=rcontrast)

names(renamed_res) <- rename_contrasts



l <- renamed_res %>%
  bind_rows() %>%
  filter(baseMean>5) %>%
  filter(cell_type %in% plot_cells, rcontrast %in% plot_contrasts) %>%
  mutate(contrast_cell_type=paste0(rcontrast," ",cell_type)) %>%
  mutate(contrast_cell_type=ifelse(grepl(" B", contrast_cell_type), paste0(contrast_cell_type," cells"),
                                   contrast_cell_type)) %>%
  split(f=.$contrast_cell_type) 

plot_order <- c("HNP1_vs_X\n_B_cells","ClY21-HNP1\n_vs_X_B_cells", "IY16-HNP1\n_vs_X_B_cells", "HNP1_vs_X\n_Mono","ClY21-HNP1\n_vs_X_Mono", "IY16-HNP1\n_vs_X_Mono","HNP1_vs_X\n_DC","ClY21-HNP1\n_vs_X_DC", "IY16-HNP1\n_vs_X_DC" ) %>% str_replace_all("_"," ")


l <- l[plot_order]

  genes <- lapply(l, function(x) x %>%
                    filter(!is.na(padj)) %>%
                    arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = msig[sel])
  
  
  
  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange,
                                                  pval = x$padj, lfc.thr=0, pval.thr = 0.05, mset = msig[sel]))
  names(sgenes) <- names(l)
  
  # s7a <- ggPanelplot(res.tmod, sgenes = sgenes, auc_thr = 0,
  #                    q_thr = 1,
  #                    filter_row_q = 1e-5,
  #                    filter_row_auc = 0.70,
  #                    q_cutoff = 1e-12,
  #                   label_angle=90,add_ids = T) +
  #   theme(axis.text=element_text(size=8),
  #         strip.text=element_text(size=8),
  #         legend.text=element_text(size=8),
  #         legend.title=element_text(size=8),
  #         axis.title=element_text(size=8),
  #         strip.text.x = element_text(hjust=0),
  #         axis.text.x=element_text(angle=0),
  #         panel.background=element_blank(),
  #         legend.key.size = unit(.1,"in"))+
  #   guides(alpha=guide_legend(ncol=1))
  # print(s7a)

ids <- c("M983", "M5928", "M42986", "M15907", "M10974", "M10214")
s7a <- res.tmod %>% 
  lapply(function(x) x %>% 
           filter(ID %in% ids)) %>%
  ggPanelplot(sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-6,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90,add_ids = T) +
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))+
    guides(alpha=guide_legend(ncol=1))
  print(s7a)
    
  
```


Per contrast.
```{r s6g_excluded, results=T, fig.width=8,fig.height=3}

data(tmod)

plot_cells <- c("B", "Mono", "CD4_T", "CD8_T", "DC")

plot_contrasts <- c("ClY21-HNP1_vs_X") %>% str_replace_all("_"," ")

rename_contrasts <- structure(c( "ClY21-HNP1\n_vs_IY16-HNP1", "ClY21-HNP1\n_vs_HNP1", "ClY21-HNP1_vs_X","IY16-HNP1\n_vs_HNP1", "IY16-HNP1\n_vs_X", "HNP1_vs_X\n" ) %>% str_replace_all("_"," "),
                              names=c("Cl_vs_I" ,"Cl_vs_UM","Cl_vs_X", "I_vs_UM","I_vs_X", "UM_vs_X"))

rcontrast <- lapply(res, function(x) rename_contrasts[x %>% pull(contrast)])
renamed_res <- Map(cbind, res, rcontrast=rcontrast)

names(renamed_res) <- rename_contrasts

l <- renamed_res %>%
  bind_rows() %>%
  filter(baseMean>5) %>%
  filter(cell_type %in% plot_cells, rcontrast %in% plot_contrasts) %>%
  split(f=.$cell_type) 
l <- l[plot_cells]

  genes <- lapply(l, function(x) x %>%
                    filter(!is.na(padj)) %>%
                    arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = tmod)
  
  
  
  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange,
                                                  pval = x$padj, lfc.thr=0, pval.thr = 0.05, mset = tmod))
  names(sgenes) <- names(l)
  
  s6g_excluded <- ggPanelplot(res.tmod, sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-5,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90, add_ids = F) + 
      ggtitle(plot_contrasts)+
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))
  print(s6g_excluded)
  

  

```


Per contrast other pathways.
```{r s7b, results=T, fig.width=8,fig.height=3}

data(tmod)

plot_cells <- c("B", "Mono", "CD4_T", "CD8_T", "DC")

plot_contrasts <- c("ClY21-HNP1_vs_X") %>% str_replace_all("_"," ")

rename_contrasts <- structure(c( "ClY21-HNP1\n_vs_IY16-HNP1", "ClY21-HNP1\n_vs_HNP1", "ClY21-HNP1_vs_X","IY16-HNP1\n_vs_HNP1", "IY16-HNP1\n_vs_X", "HNP1_vs_X\n" ) %>% str_replace_all("_"," "),
                              names=c("Cl_vs_I" ,"Cl_vs_UM","Cl_vs_X", "I_vs_UM","I_vs_X", "UM_vs_X"))

rcontrast <- lapply(res, function(x) rename_contrasts[x %>% pull(contrast)])
renamed_res <- Map(cbind, res, rcontrast=rcontrast)

names(renamed_res) <- rename_contrasts


l <- renamed_res %>%
  bind_rows() %>%
  filter(baseMean>5) %>%
  filter(cell_type %in% plot_cells, rcontrast %in% plot_contrasts) %>%
  split(f=.$cell_type) 
l <- l[plot_cells]

  genes <- lapply(l, function(x) x %>%
                    filter(!is.na(padj)) %>%
                    arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = msig[sel])
  
  
  
  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange,
                                                  pval = x$padj, lfc.thr=0, pval.thr = 0.05, mset = msig[sel]))
  names(sgenes) <- names(l)
  
  s7b <- ggPanelplot(res.tmod, sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-5,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90, add_ids=T) + 
      ggtitle(plot_contrasts)+
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))
  print(s7b)
  
  
ids <- c("M973", "M5928", "M16031","M15907", "M10974")
s7b <- res.tmod %>% 
  lapply(function(x) x %>% 
           filter(ID %in% ids)) %>%
  ggPanelplot(sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-5,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90,add_ids = F) +
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))+
    guides(alpha=guide_legend(ncol=1))
print(s7b)


```

```{r s7c, results=T, fig.width=8,fig.height=3}

data(tmod)

plot_cells <- c("B", "Mono", "CD4_T", "CD8_T", "DC")

plot_contrasts <- c("IY16-HNP1_vs_X") %>% str_replace_all("_"," ")

rename_contrasts <- structure(c( "ClY21-HNP1\n_vs_IY16-HNP1", "ClY21-HNP1\n_vs_HNP1", "ClY21-HNP1_vs_X","IY16-HNP1\n_vs_HNP1", "IY16-HNP1_vs_X", "HNP1_vs_X\n" ) %>% str_replace_all("_"," "),
                              names=c("Cl_vs_I" ,"Cl_vs_UM","Cl_vs_X", "I_vs_UM","I_vs_X", "UM_vs_X"))

rcontrast <- lapply(res, function(x) rename_contrasts[x %>% pull(contrast)])
renamed_res <- Map(cbind, res, rcontrast=rcontrast)

names(renamed_res) <- rename_contrasts


l <- renamed_res %>%
  bind_rows() %>%
  filter(baseMean>5) %>%
  filter(cell_type %in% plot_cells, rcontrast %in% plot_contrasts) %>%
  split(f=.$cell_type) 
l <- l[plot_cells]

  genes <- lapply(l, function(x) x %>%
                    filter(!is.na(padj)) %>%
                    arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = msig[sel])
  
  
  
  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange,
                                                  pval = x$padj, lfc.thr=0, pval.thr = 0.05, mset = msig[sel]))
  names(sgenes) <- names(l)
  
  s7c <- ggPanelplot(res.tmod, sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-7,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90, add_ids=T) + 
      ggtitle(plot_contrasts)+
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))
  print(s7c)
  

ids <- c("M983", "M42986", "M42828","M15907", "M11125","M10974")
s7c <- res.tmod %>% 
  lapply(function(x) x %>% 
           filter(ID %in% ids)) %>%
  ggPanelplot(sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-5,
                     filter_row_auc = 0.70,
                     q_cutoff = 1e-12,
                    label_angle=90,add_ids = T) +
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          axis.text.x=element_text(angle=0),
          panel.background=element_blank(),
          legend.key.size = unit(.1,"in"))+
    guides(alpha=guide_legend(ncol=1))
print(s7c)

```


```{r tmod3, results=T, fig.width=15,fig.height=10}

data(tmod)

contrasts <- names(res)

plot_panels <- function(contrast){
  # contrast <- contrasts[1]
  l <- split(res[[contrast]], f=res[[contrast]]$cell_type)
  genes <- lapply(l, function(x) x %>%
                    filter(!is.na(padj)) %>%
                    arrange(pvalue))
  lgenes <- lapply(genes, function(x) x$gene_name)
  res.tmod <- lapply(lgenes, tmodCERNOtest, mset = tmod)
  
  
  sgenes <- sapply(l, function(x) tmodDecideTests(x$gene_name, lfc = x$log2FoldChange,
                                                  pval = x$padj, lfc.thr=0, pval.thr = 0.05, mset = tmod))
  names(sgenes) <- names(l)
  
  
  keep_results <- names(res.tmod)[sapply(res.tmod, function(x) x %>% filter(adj.P.Val<1e-4 , AUC >0.65)%>% nrow() >0)]
  relevant_results <- res.tmod[keep_results]
  cat_cell_type <- lapply(names(relevant_results), function(x) rep(x, nrow(relevant_results[[x]])))
  relevant_results <- Map(cbind, relevant_results, cell_type=cat_cell_type)
  
  if(length(relevant_results)>0){
    sgenes <- sgenes[names(relevant_results)]
    p <- ggPanelplot(relevant_results, sgenes = sgenes, auc_thr = 0,
                     q_thr = 1,
                     filter_row_q = 1e-4,
                     filter_row_auc = 0.65,
                     q_cutoff = 1e-12) + 
      ggtitle(contrast)
    print(p)
  }
  
  relevant_results <- do.call(rbind, relevant_results)
  return(relevant_results)
}

res_tmod <- lapply(contrasts, plot_panels)
names(res_tmod) <- contrasts
res_tmod <- res_tmod[!sapply(res_tmod, function(x) length(x)==0)]

```


```{r f5f f5g, fig.width=4,fig.height=3}

chemokine_modules <- c("LI.M27.1",
                       "LI.M59")

contrast <- "Cl_vs_X"
module <- "LI.M27.1"
module_full <- tmod$gs[tmod$gs$ID==module,"Title"]

cell_type <- "B"

l <- split(res[[contrast]], f=res[[contrast]]$cell_type)[cell_type]

genes <- lapply(l, function(x) x %>%
                  filter(!is.na(padj)) %>%
                  arrange(pvalue))
lgenes <- lapply(genes, function(x) x$gene_name)[[cell_type]]

cols <- ifelse(genes[[cell_type]]$log2FoldChange >0 & genes[[cell_type]]$padj <0.05, "red",
                  ifelse(genes[[cell_type]]$log2FoldChange <0 & genes[[cell_type]]$padj <0.05, "blue",
                         ifelse(genes[[cell_type]]$padj>0.05, "grey", NA)))
names(cols) <- lgenes

tmod_pvalue <- res_tmod[[contrast]] %>%
  filter(cell_type==cell_type) %>%
  filter(ID==module) %>%
  pull(adj.P.Val) %>%
  signif(digits = 3)

update_geom_defaults("text_repel", list(size = 2.5))
f5f <- ggEvidencePlot(l = lgenes, m=module, mset = tmod, gene.colors = cols ) +
  ggtitle(paste0(module_full,"\n","adj.P.val=",tmod_pvalue,"\n","ClY21-HNP1 vs X","\n",cell_type, " cells"))+
  theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8))

f5f


contrast <- "Cl_vs_X"
module <- "LI.M59"
module_full <- tmod$gs[tmod$gs$ID==module,"Title"]

cell_type <- "B"

l <- split(res[[contrast]], f=res[[contrast]]$cell_type)[cell_type]

genes <- lapply(l, function(x) x %>%
                  filter(!is.na(padj)) %>%
                  arrange(pvalue))
lgenes <- lapply(genes, function(x) x$gene_name)[[cell_type]]

cols <- ifelse(genes[[cell_type]]$log2FoldChange >0 & genes[[cell_type]]$padj <0.05, "red",
                  ifelse(genes[[cell_type]]$log2FoldChange <0 & genes[[cell_type]]$padj <0.05, "blue",
                         ifelse(genes[[cell_type]]$padj>0.05, "grey", NA)))
names(cols) <- lgenes

tmod_pvalue <- res_tmod[[contrast]] %>%
  filter(cell_type==cell_type) %>%
  filter(ID==module) %>%
  pull(adj.P.Val) %>%
  signif(digits = 3)

update_geom_defaults("text_repel", list(size = 2.5))
f5g <- ggEvidencePlot(l = lgenes, m=module, mset = tmod, gene.colors = cols ) +
  ggtitle(paste0(module_full,"\n","adj.P.val=",tmod_pvalue,"\n","ClY21-HNP1 vs X","\n",cell_type, " cells"))+
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8))

f5g


```



```{r s7d s7e, fig.width=4,fig.height=3}

contrast <- "I_vs_X"
module <- "LI.M27.1"
module_full <- tmod$gs[tmod$gs$ID==module,"Title"]

cell_type <- "Mono"

l <- split(res[[contrast]], f=res[[contrast]]$cell_type)[cell_type]

genes <- lapply(l, function(x) x %>%
                  filter(!is.na(padj)) %>%
                  arrange(pvalue))
lgenes <- lapply(genes, function(x) x$gene_name)[[cell_type]]

cols <- ifelse(genes[[cell_type]]$log2FoldChange >0 & genes[[cell_type]]$padj <0.05, "red",
                  ifelse(genes[[cell_type]]$log2FoldChange <0 & genes[[cell_type]]$padj <0.05, "blue",
                         ifelse(genes[[cell_type]]$padj>0.05, "grey", NA)))
names(cols) <- lgenes

tmod_pvalue <- res_tmod[[contrast]] %>%
  filter(cell_type==cell_type) %>%
  filter(ID==module) %>%
  pull(adj.P.Val) %>%
  signif(digits = 3)

update_geom_defaults("text_repel", list(size = 2.5))
s7d <- ggEvidencePlot(l = lgenes, m=module, mset = tmod, gene.colors = cols ) +
  ggtitle(paste0(module_full,"\n","adj.P.val=",tmod_pvalue,"\n","IY16-HNP1 vs X","\n",cell_type, "cytes"))+
  theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8))

s7d


contrast <- "I_vs_X"
module <- "DC.M5.12"
module_full <- tmod$gs[tmod$gs$ID==module,"Title"]

cell_type <- "Mono"

l <- split(res[[contrast]], f=res[[contrast]]$cell_type)[cell_type]

genes <- lapply(l, function(x) x %>%
                  filter(!is.na(padj)) %>%
                  arrange(pvalue))
lgenes <- lapply(genes, function(x) x$gene_name)[[cell_type]]

cols <- ifelse(genes[[cell_type]]$log2FoldChange >0 & genes[[cell_type]]$padj <0.05, "red",
                  ifelse(genes[[cell_type]]$log2FoldChange <0 & genes[[cell_type]]$padj <0.05, "blue",
                         ifelse(genes[[cell_type]]$padj>0.05, "grey", NA)))
names(cols) <- lgenes

tmod_pvalue <- res_tmod[[contrast]] %>%
  filter(cell_type==cell_type) %>%
  filter(ID==module) %>%
  pull(adj.P.Val) %>%
  signif(digits = 3)

update_geom_defaults("text_repel", list(size = 2.5))
s7e <- ggEvidencePlot(l = lgenes, m=module, mset = tmod, gene.colors = cols ) +
  ggtitle(paste0(module_full,"\n","adj.P.val=",tmod_pvalue,"\n","IY16-HNP1 vs X","\n",cell_type, "cytes"))+
    theme(axis.text=element_text(size=8),
          strip.text=element_text(size=8),
          legend.text=element_text(size=8),
          legend.title=element_text(size=8),
          axis.title=element_text(size=8),
          strip.text.x = element_text(hjust=0),
          plot.title=element_text(size=8))

s7e


```


```{r fig5,fig.width=8,fig.height=12}

(f5a + plot_spacer()) /
  wrap_elements(f5b) /
  (((f5c | f5d |f5e ) +plot_layout(guides="collect")))/
  (f5f|f5g)+
  plot_layout(heights = c(.8,1.2,.8,.7)) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 8, face = "bold",hjust = -.5, vjust = 0))

```

```{r save}

ggsave(file.path(path,results_dir,"fig5.pdf"), device = cairo_pdf,width=8,height=12)

```


Supplement: Show cell type proportion (l1).

```{r prop,fig.width=8,fig.height=3}

my_comparisons <- list(c("UM","X"),
                       c("Cl","X"),
                       c("I","X"))

plot_groups <- unlist(my_comparisons) %>% unique

plot_cells <- c("B", "CD4 T", "CD8 T", "Mono", "DC")


s6b <- sobj[[]] %>%
  select(sample,group,predicted.celltype.l1) %>%
  group_by(sample,group,predicted.celltype.l1) %>%
  summarise(n=n()) %>%
  mutate(tot_cells=sum(n),pct.cells=n/sum(n)*100) %>%
  ungroup() %>%
  filter(group %in% plot_groups, predicted.celltype.l1 %in% plot_cells) %>%
  ggplot(aes(group,pct.cells,colour=group))+
  geom_point(position=position_jitterdodge(jitter.width = .3),shape=1)+
  #stat_compare_means(method="t.test", paired=F,comparisons = my_comparisons,label = "p.format",size=3)+
  facet_wrap(~predicted.celltype.l1, strip.position = "top",nrow=1)+
  scale_colour_manual(labels=c("ClY21-HNP1", "IY16-HNP1", "HNP1", "X" ), values=c("dodgerblue3","orange2", "darkgreen","grey20"))+
  theme_classic()+
  theme(axis.text.x=element_blank(),
        strip.background = element_blank(),
        legend.title=element_blank(),
        strip.text=element_text(size=8),
        axis.text=element_text(size=8),
        axis.title=element_text(size=8),
        legend.text=element_text(size=8),
        legend.position = "bottom")+
  labs(y=c("% cells"),x=c())
s6b

```

```{r s6a}

md <- sobj[[]] %>%
  select(sample,nCount_RNA ,nFeature_RNA,predicted.celltype.l1.score,predicted.celltype.l2.score) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(sample=paste0(sample,' (n=',n(),')')) %>%
  gather(metric,value,-sample)

s6a <- ggplot(md,aes(x=sample,y=value,fill=sample)) + 
  geom_violin(scale='width') +
  facet_wrap(~metric,scales='free_y',ncol=2) + 
  theme_classic() + 
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
        legend.position='none',
        strip.background = element_blank(),
        strip.text = element_text(size=8),
        axis.text=element_text(size=8)) + 
  labs(x='',y='')

s6a

```


```{r sfig6,fig.width=8,fig.height=12}

 wrap_elements(s6a) / 
  wrap_elements(s6b) /
  (wrap_elements(s6c)|wrap_elements(s6d)) /
  (wrap_elements((s6e|s6f)+plot_layout(guides="collect"))) +
  plot_layout(heights = c(1.4,1,1,1)) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 8, face = "bold",hjust = -.5, vjust = 0))


```

```{r save supp 1}

ggsave(file.path(path,results_dir,"Sfig6.pdf"), device = cairo_pdf,width=8,height=12)
```

```{r sfig7,fig.width=8,fig.height=12}

wrap_elements(s7a) / 
  wrap_elements(s7b) /
  wrap_elements(s7c) /
  (s7d + s7e) +
  plot_layout(heights = c(1.2,1.2,1.2,.8)) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 8, face = "bold",hjust = -.5, vjust = 0))


```

```{r save supp 2}

ggsave(file.path(path,results_dir,"Sfig7.pdf"), device = cairo_pdf,width=8,height=12)
```


```{r sfig8,fig.width=8,fig.height=3}

 (wrap_elements((s8a | s8b |s8c ) +plot_layout(guides="collect")))+
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 8, face = "bold",hjust = -.5, vjust = 0))


```


```{r save supp 3}

ggsave(file.path(path,results_dir,"Sfig8.pdf"), device = cairo_pdf,width=8,height=3)


```





















```{r plot chemokine genes}

chemokine_modules <- c("LI.M27.1",
                       "LI.M59")

contrast <- "Cl_vs_X"
module <- "LI.M27.1"
module_full <- tmod$gs[tmod$gs$ID==module,"Title"]

cell_type <- "B"

genes_in_gene_set <- 
  tmod$gv[tmod$gs2gv[[which(tmod$gs$ID==module)]]]



# df_colors <- res[[contrast]] %>% 
#   filter(gene_name %in% genes_in_gene_set, cell_type==cell_type) %>%
#   mutate(color=ifelse(log2FoldChange >0 & padj < .05, "red",
#                       ifelse(log2FoldChange <0 & padj < .05, "blue",
#                              "grey"))) %>%
#   select(gene_name,color)
# 
# gene.colors <- df_colors$color
# names(gene.colors) <- df_colors$gene_name




p10 <- res[[contrast]] %>%                  
    dplyr::filter(!is.na(padj), baseMean>10, cell_type=="B") %>%
    ggplot(aes(x=log2FoldChange,y=-log10(pvalue),color=padj < .05)) + 
    geom_point(size=.5, shape=1) +
    geom_label_repel(data=res[[contrast]] %>% 
                       dplyr::filter(!is.na(padj), baseMean>10, cell_type=="B",gene_name %in% genes_in_gene_set),
                       aes(label=gene_name),
                     segment.size=.25,segment.alpha=.5,size=3,color='black',max.overlaps=30,
                     label.padding=0.05,label.size=NA) +
    scale_color_manual(values=c('black','red')) +
    theme_classic() +
    theme(legend.position='none',
          strip.background=element_blank())+
    ggtitle(paste0(contrast, " B cells"))
print(p10)


contrast <- "I_vs_X"
p11 <- res[[contrast]] %>%                  
    dplyr::filter(!is.na(padj), baseMean>10, cell_type=="B") %>%
    ggplot(aes(x=log2FoldChange,y=-log10(pvalue),color=padj < .05)) + 
    geom_point(size=.5, shape=1) +
    geom_label_repel(data=res[[contrast]] %>% 
                       dplyr::filter(!is.na(padj), baseMean>10, cell_type=="B",gene_name %in% genes_in_gene_set),
                       aes(label=gene_name),
                     segment.size=.25,segment.alpha=.5,size=3,color='black',max.overlaps=30,
                     label.padding=0.05,label.size=NA) +
    scale_color_manual(values=c('black','red')) +
    theme_classic() +
    theme(legend.position='none',
          strip.background=element_blank())+
    ggtitle(paste0(contrast, " B cells"))
print(p11)

contrast <- "UM_vs_X"
p12<- res[[contrast]] %>%                  
    dplyr::filter(!is.na(padj), baseMean>10, cell_type=="B") %>%
    ggplot(aes(x=log2FoldChange,y=-log10(pvalue),color=padj < .05)) + 
    geom_point(size=.5, shape=1) +
    geom_label_repel(data=res[[contrast]] %>% 
                       dplyr::filter(!is.na(padj), baseMean>10, cell_type=="B",gene_name %in% genes_in_gene_set),
                       aes(label=gene_name),
                     segment.size=.25,segment.alpha=.5,size=3,color='black',max.overlaps=30,
                     label.padding=0.05,label.size=NA) +
    scale_color_manual(values=c('black','red')) +
    theme_classic() +
    theme(legend.position='none',
          strip.background=element_blank())+
    ggtitle(paste0(contrast, " B cells"))
print(p12)

```

concordance disco plot

```{r conc}

cont1 <- "Cl_vs_X"
cont2 <- "I_vs_X"


df <- data.frame(gene_name=res[[cont1]]$gene_name,
                 log2FoldChange_cont1=res[[cont1]]$log2FoldChange,
                 padj_cont1=res[[cont1]]$padj,
                 log2FoldChange_cont2=res[[cont2]]$log2FoldChange,
                 padj_cont2=res[[cont2]]$padj,
                 cell_type=res[[cont1]]$cell_type)

df <- df %>%
  mutate(regulation1=ifelse(log2FoldChange_cont1>0 & padj_cont1 < 0.05 , "up_cont1",
                            ifelse(log2FoldChange_cont1<0 & padj_cont1 < 0.05 , "down_cont1",NA)),
         regulation2=ifelse(log2FoldChange_cont2>0 & padj_cont2 < 0.05 , "up_cont2",
                            ifelse(log2FoldChange_cont2<0& padj_cont2 < 0.05 , "down_cont2",NA)),
         regulation=ifelse(regulation1=="up_cont1" & regulation2=="down_cont2", "up_down",
                           ifelse(regulation1=="up_cont1" & regulation2=="up_cont2", "up_up",
                                  ifelse(regulation1=="down_cont1" & regulation2=="up_cont2", "down_up",
                                         ifelse(regulation1=="down_cont1" & regulation2=="down_cont2", "down_down",NA)))))
                         


df %>% 
  filter(cell_type=="B") %>%
  filter(!is.na(padj_cont1) | !is.na(padj_cont2) ) %>%
  ggplot(aes(log2FoldChange_cont1 ,log2FoldChange_cont2, color=regulation ))+
  geom_abline(slope = 1,lty=2)+
  geom_point(size=1,shape=1)+
  geom_label_repel(data=df %>% 
                     filter(cell_type=="B",gene_name %in% genes_in_gene_set) %>%
                     filter(!is.na(padj_cont1) | !is.na(padj_cont2) ),
                   aes(label=gene_name),
                   segment.size=.25,segment.alpha=.5,size=3,color='black',max.overlaps=100,
                   label.padding=0.05,label.size=NA) +
  theme_classic() +
  theme(strip.background=element_blank())
```





```{r ma, fig.height=25, fig.width=10, eval=F}

for (comparison in unique(res_df$contrast)){
  p <- res_df %>%
    dplyr::filter(contrast == comparison) %>%
    ggplot(aes(x=baseMean,y=log2FoldChange,color=padj < .05)) +
    geom_point(size=.5, shape=1) +
    geom_label_repel(data=res_df %>%
                       dplyr::filter(contrast == comparison, padj < .05),
                     aes(label=gene_name),
                     segment.size=.25,segment.alpha=.5,size=3,color='black',max.overlaps=50,
                     label.padding=0.05,label.size=NA) +
    scale_color_manual(values=c('black','red')) +
    scale_x_continuous(trans='log10') +
    coord_cartesian(ylim =c(-3,3)) +
    theme_classic() +
    theme(legend.position='none',
          strip.background=element_blank())+
    xlab("mean normalized counts")+
    ggtitle(comparison)+
    facet_wrap(~cell_type, ncol = 2)
  print(p)
}
``` 





```{r session}

sessionInfo()

```


